Spike-and-Slab Generalized Additive Models and Scalable Algorithms for High-Dimensional Data

### 2021-09-26 Manuscript/02-Method.Rmd

Following the GLM notation introduced in Section \ref{sec:intro}, we have a generalized additive model and its matrix form,
\begin{equation}\label{eq:gam}
E(Y|\bs x) = g^{-1}(a + \sum\limits^p_{j=1}B_j(x_j)) = g^{-1}(a + \sum\limits^p_{j=1} \bs \beta_j^T \bs X_j),
\end{equation}
and the data distribution is
\begin{equation}
f(y|X\beta, \phi) = \prod\limits^n_{i=1}f(y_i|X_i,\phi),\nonumber
\end{equation}
with smoothing functions $B_j(x_j)$ of the variable $x_j, j = 1, \dots, p.$ The basis function matrix, i.e. the design matrix derived from the smoothing function $B_j(x_j)$, is denoted $\bs X_j$ for the variable $x_j$. The dimension of the design matrix depends on the choice of the smoothing function. For example, a cubic spline function with $k$ knots produces a $k+3$ elements column vector. Its corresponding smoothing penalty matrix is denoted as $\bs S_j$. $\bs \beta_j$ denotes the basis function coefficients for the $j$th variable, and the matrix transposing operation is denoted with a superscript $^T$. To note, the proposed model can include parametric forms of variables in the model, treating the parametric function as a special case of the smoothing function, e.g. $B_j(x_j) = x_j$ with the smoothing penalty matrix defined as $\bs S_j = \begin{bmatrix}0\end{bmatrix}$.

To encourage proper smoothing of the functions, we adopt the idea of smoothing penalties from smoothing spline models. The basic idea is to set up a smoothing penalty, described in Equation (\ref{eq:smoothpen}), in the prior density function, or least squared and likelihood function in the penalized framework. However, the direct integration of smoothing penalties into sparsity penalties is not obvious. Marra and  Wood [@Marra2011] proposed a re-parameterization procedure to accommodate the smoothing penalty implicitly. Given the smoothing penalty matrix is most often symmetric and positive semi-definite for univariate smoothing functions, we apply eigendecomposition on the penalty matrix $\bs S = \bs U \bs D \bs U^T$ , where the matrix $\bs D$ is diagonal with the eigenvalues arranged in the ascending order. To note, $\bs D$ can contain $0$ elements on the diagonal, associated with the linear space of the smoothing function. We further write the orthonormal matrix $\bs U \equiv \begin{bmatrix}\bs U^0 : \bs U^{\tp}\end{bmatrix}$ containing the eigenvectors as columns in the corresponding order. That is $\bs U^0$ contains the eigenvectors with zero eigenvalues for the linear space and $\bs U^{\tp}$ contains the eigenvectors (as columns) for the non-zero eigenvalues, i.e. the non-linear space. We multiply the basis function matrix $\bs X$ with the orthonormal matrix $\bs U$ for the new design matrix $\bs X^\ast = \bs X \bs U \equiv \begin{bmatrix} \bs X^0 : \bs X^{\tp} \end{bmatrix}$. An additional scaling step is imposed on $\bs X^{\tp}$ by the non-zero eigenvalues of $\bs D$ such that the new basis function matrix $\bs X^\ast$ can receive uniform penalty on each of its dimensions. With slight abuse of the notation, we denote $\bs X_j \equiv \begin{bmatrix} \bs X_j^0 : \bs X_j^{\tp} \end{bmatrix}$ as the basis function matrix for the $j$th variable after the re-parameterization. A spline function can be expressed in the matrix form
$$
B_j(x_j) = B_j^0(x_j) + B_j^\tp (x_j) = \bs {\beta_j^0}^T \bs X_j^0 + \bs {\beta_j^\tp}^T \bs X_j^\tp ,
$$
and the generalized additive model in Equation (\ref{eq:gam}) be
\begin{equation}\label{eq:gam-repa}
E(Y|\bs x) =  g^{-1}(a + \sum\limits^p_{j=1} \bs \beta_j^T \bs X_j) = g^{-1}\left[a + \sum\limits^p_{j=1} (\bs {\beta_j^0}^T \bs X_j^0 + {\bs \beta_j^\tp}^T \bs X_j^\tp)\right].
\end{equation}

@@

The major criticism of early spike-and-slab priors receive is being computationally prohibitive. [@Bai2021Review] Since then, many studies focus on alleviating the computational burden that sampling algorithms bear, which include EMVS [@Rockova2014a] and Spike-and-slab Lasso [@Rockova2018b; @Rockova2018]. Particularly, the development of Spike-and-slab Lasso  substantially improves the scalability of SS models, setting up the theoretical foundation for generalized models in -omics data analysis [@Tang2017a; @Tang2017; @Tang2018; @Tang2019]. The spike-and-slab lasso prior is composed of two double exponential distributions with mean 0 and different dispersion parameters, $0 < s_0 < s_1$, mathematically,
\begin{equation} 
\beta | \gamma \sim (1-\gamma)DE(0, s_0) + \gamma DE(0, s_1), 0 < s_0 < s_1.\nonumber
\end{equation}
Given that both double exponential distributions have a mean of 0 and the latent indicator $\gamma$ can only take the value of 0 or 1, the mixture double exponential distribution can be formulated as a double exponential density,
\begin{equation} \label{eq:ssl}
\beta | \gamma \sim DE(0, (1-\gamma^0_{j}) s_0 + \gamma^0_{j} s_1), 0 < s_0 < s_1,
\end{equation}
with the scale parameter $S = (1-\gamma)s_0 + \gamma s_1$. The SSL also mitigates the problem of EMVS where the weak signals are not shrink to zero, and hence is preferred in high-dimensional data analysis. We notice that Bai [@Bai2021] is the first to apply spike-and-slab lasso prior in the GAM framework, where the densities of the spike and slab components take the group lasso density [@Xu2015]. Bai [@Bai2021] takes an "all-in-all-out" strategy for function selection.

### Spike-and-Slab Lasso Prior
We introduce a novel prior for GAMs, particularly for high-dimensional nonlinear modeling with bi-level selection. The proposed prior extends from the spike-and-slab lasso prior described in Equation (\ref{eq:ssl}). Given the re-parameterized deign matrix $\bs X_j = \begin{bmatrix} \bs X_j^0 : \bs X_j^\tp \end{bmatrix}$ for the $j$th variable, we impose a two-part SSL prior to the coefficients $\bs \beta_j$. Specifically, we impose independent group priors on the linear space coefficients and on the nonlinear space coefficients respectively,

\begin{align}\label{eq:bham_ssl}
  \beta^0_{j} |\gamma^0_{j},s_0,s_1 &\sim DE(0,(1-\gamma^0_{j}) s_0 + \gamma^0_{j} s_1) \nonumber \\
  \beta^\tp_{jk} | \gamma^\tp_{j},s_0,s_1 &\sim DE(0,(1-\gamma^\tp_{j}) s_0 + \gamma^\tp_{j} s_1), 
\end{align}

where $\gamma^0_{j}\in\{0,1\}$ and $\gamma^\tp_{j}\in \{0,1\}^{K_j}$ are latent indicator variable and vectors, indicating if the model includes the linear effect and the nonlinear effect of the $j$th variable respectively. $s_0$ and $s_1$ are scale parameters, assuming $0 < s_0 < s_1$ and given. These scale parameters $s_0$ and $s_1$ can be treated as tuning parameters and optimized via cross-validation. A discussion of how to choose the scale parameters comes in Section \ref{sec:tune}. To note, this prior differs from previous group spike-and-slab lasso priors [@Tang2018; @Tang2019], where the $\bs \beta^0_j$ and $\bs \beta^\tp_j$ shares the same indicator variables $\gamma_j^0$, $\gamma_j^\tp$ respectively. It is possible to add a more restrictive assumption on the priors, assuming that one indicator variable decides the inclusion of both the linear effect and nonlinear effect, i.e. $\gamma_j = \gamma^0_j = \gamma^\tp_j$.  This converges to the SB-GAM [@Bai2021]. Conversely, it is also possible to relax the assumption such that each coefficient $\beta$ to have its own latent indicator $\gamma$, but at the cost of complicating the bi-level function selection. This reduces the proposed prior to the classic spike-and-slab lasso prior.

@@

The rest of the hierarchical prior follows the traditional spike-and-slab lasso prior: we set up hyper-priors of $\bs \gamma$ to allow local adaption of the shrinkage using a Bernoulli distribution (binomial distribution of one trial). The two indicators of the $j$th predictor, $\gamma^{0}_j$ and $\gamma^\tp_j$, shares the same probability parameter $\theta_j$,
$$
\begin{aligned}
\gamma_{j}^{0} | \theta_j &\sim Bin(\gamma^{0}_{j}|1, \theta_j)\\
\gamma_{j}^\tp | \theta_j &\sim Bin(\gamma^\tp_{j}|1, \theta_j).
\end{aligned}
$$
This is to leverage the fact that the probability of selecting the bases of a smoothing function should be similar, while allowing different penalty on the linear space and non-linear space of the smoothing function. The hyper prior of $\gamma_{j}^{0}$ decides the sparsity of the model at the function selection level, while that of $\gamma_{j}^\tp$ decides the smoothness of the spline function at basis function level. Meanwhile, we specify that $\gamma_{j}^0$ and $\gamma_{j}^\tp$ are independently distributed for analytic simplicity. We further specify the parameter $\theta_j$ follows a beta distribution with given shape parameters $a$ and $b$,
$$
\theta_j \sim Beta(a, b).
$$
The beta distribution is a conjugate prior for the binomial distribution and hence provides some computation convenience. For simplicity, we focus on a special case of beta distribution, uniform (0,1). When the variable have large effects in any of the bases, the parameter $\theta_j$ will be estimated large, which in turn encourages the model to include the rest of bases, achieving the local adaption among spline bases. Hereafter, we refer Bayesian hierarchical generalized additive models with the spike-and-slab lasso prior as the BHAM-SSL, and visually presented in Figure \ref{fig:SSprior}.

@@

\matrix (linearPrior) [matrix of nodes, column sep = 0mm, row sep = 0.7cm] {
  \node (linearGamma) [outCome] { $\gamma_j^0 \sim Bin(1, \theta_j) $ };\\
  \node (linearBeta) [outCome] { $\beta_j^0 \sim DE(0,(1-\gamma^0_{j}) s_0 + \gamma^0_{j} s_1)$};\\
};
\matrix (penPrior) [right = 2cm of linearPrior, matrix of nodes, column sep = 0mm, row sep = 0.7cm] {
  \node (penGamma) [outCome] { $\gamma_{j}^\tp \sim Bin(1, \theta_j)$ };\\
  \node (penBeta) [outCome] { $\beta_{jk}^\tp \sim  DE(0,(1-\gamma^\tp_{j}) s_0 + \gamma^\tp_{j} s_1)$};\\
};

\node (s) [staticCompo]  at ($(linearBeta)!0.5!(penBeta)$)  {($s_0, s_1$)};
\node (Beta) [staticCompo, below = 1cm of s] {$\bs \beta = (\beta^0_1, \bs \beta^\tp_1, \dots,\beta^0_j, \bs \beta^\tp_j , \dots,\beta^0_p, \bs \beta^\tp_p) $};
\node (Theta)[outCome, above = 2cm of s] {$\theta_{j} \sim Beta(a, b)$};
\node (ab)[staticCompo, above = 0.5cm of Theta] {$(a, b)$};
\node (Y) [outCome, below = 1cm of Beta] {$y_i \sim Expo. Fam. (g^{-1}(\bs \beta^T \bs X_i), \phi)$};

@@

With the re-parameterization step of the basis function matrix $X$, it is possible to generalized the SSL prior to other priors, for example normal priors for ridge-type regularization and mixture normal prior for spike-and-slab regularization. These priors would work better in low and medium dimensional settings where the sparse assumption is not necessary. Here we focus on the mixture normal prior as a demonstration of applying continuous spike-and-slab prior in additive models.

A spike-and-slab mixture normal prior in additive models can be expressed as 
$$
\begin{aligned}
  \beta^0_{j} |\gamma^0_{j},s_0,s_1 &\sim N(0,(1-\gamma^0_{j}) s_0 + \gamma^0_{j} s_1)\nonumber\\
  \beta^\tp_{jk} | \gamma^\tp_{j},s_0,s_1 &\sim N(0,(1-\gamma^\tp_{j}) s_0 + \gamma^\tp_{j} s_1).
\end{aligned}
$$
Similar to the BHAM-SSL prior in Equation (\ref{eq:bham_ssl}), $0 < s_0 < s_1$ are tuning parameters and can be optimized via cross-validation. One of the critics received by the spike-and-slab mixture normal prior is that the tails of a normal distribution diminishes to zero too fast, which renders problem when estimating the large effects. Distributions with heavier tails can be used as an alternative, for example mixture $t$ distribution with small degree of freedom.

@@

The proposed models can be fitted with MCMC algorithms. Nevertheless, the computational burden of MCMC algorithms creates scalability issues. George and McCulloch[@George1997] examined the computation speed for various MCMC algorithms with continous spike-and-slab priors, and suggested MCMC algorithms works well for medium size ($p$=25) of predictors with only linear effects . However, it is not feasible to high-dimensional data analysis where the number of predictors easily exceeds 100. Specific to additive models, each predictor would expand to multiple new "predictors" via basis functions, creating greater computational demands. Scheipl et al. [@Scheipl2013] demonstrated the computational demands of a MCMC algorithm for fitting spike-and-slab additive models grow exponentially as $p$ increases modestly via monte carlo simulation studies. Hence, we feel compelled to develop scalable algorithms for fitting Bayesian hierarchical additive models in high-dimensional settings.

As alternatives to the sampling algorithms for fitting Bayesian models, optimization algorithms focus on the maximum a posteriori (MAP) estimates and speed up the model fitting process at the cost of uncertainty measures. The earlier work for fitting spike-and-slab models using optimization algorithms includes EMVS [@Rockova2014a]. Rockova and George[@Rockova2014a] proposed an expectation-maximization (EM) based algorithm to fit models that use continuous mixture normal priors. 
<!-- The coefficient $\beta_j$ independently follows a mixture normal distribution $N(0, \sigma^2((1-\gamma_i)v_0 + \gamma_i v_1)$ for $0 \leq v_0 < v_1$. The parameter $\sigma^2$ follows a inverse gamma prior, $IG(v/2, v\lambda /2)$. Like in the spike-and slab family, the latent binary variables $\bs \gamma$ follows a Bernoulli prior with a hyper prior $\theta$ follows a beta distribution, or simpler the uniform (0,1).  -->
In the E step, the latent binary indicators $\gamma$s are treated as the missing data, and the posterior means are calculated conditioning on the current value of other parameters; in the M step, a ridge estimator was used to update the coefficients, and $\sigma^2, \theta$ are updated accordingly. 
<!-- As the ridge regression creates small but nonzero coefficients, a further step is needed to select the variable: EMVS further suggests to threshold the inclusion probability for variable selection or generate regularization plot with different values of the spike scale parameter. EMVS is also compatible with structured prior when dealing with grouped variables. The group structure can be imposed on the prior of the latent binary variable $\gamma$ via either logistic regression or markov random field.  -->
The same authors [@Rockova2018b; @Rockova2018] further combined the EM algorithm with coordinate descent algorithm for fitting SSL models.
<!-- the continuous spike-and-slab Gaussian prior to the spike-and-slab double exponential distribution, as known as spike-and-slab lasso (SSL). The authors also lay the theoretical discussion of comparing spike-and-slab from the penalty perspective and the Bayesian perspective. The SSL also mitigated the problem of EMVS that the weak signals are not shrink to zero. -->
Yi and his group independently developed the EM-Iterative Weighted Least Square and EM-cyclic coordinate descent algorithms to fit models with broader class of priors, SSL included.[@Yi2019] These algorithms were implemented for generalized linear models[@Tang2017a], Cox proportional hazards models [@Tang2017] and their grouped counterparts[@Tang2018; @Tang2019]. Both EM based algorithms provide deterministic solutions, which becomes a more popular property for reproducible research

@@

In the follow section, we expand the two EM-based algorithms, EM-CD and EM-IWLS algorithms, to fit BHAMs. To note, the two proposed algorithms provides different utilities. The EM-CD algorithm is specifically for fitting BHAM-SSL with an expedited performance, recommending to use in high and ultra-high dimensional setting. SB-GAM[@Bai2020; @Bai2021] also used EM-CD algorithm. The main difference between proposed EM-CD algorithm and SB-GAM is that SB-GAM uses a blocked CD algorithm for their group prior. In comparison, the proposed prior is pairwise independent requiring no special treatment. A specific concern of EM-CD algorithm is that it provides no information for inference.  In contrast, the EM-IWLS can extract the variance-covariance matrix thanks to the IWLS algorithm and construct Wald type inference. Moreover, the EM-IWLS is a more general model fitting algorithm that can be used for fitting not only SSL and continuous SS priors but also Student's t-priors and double exponential priors.

@@

EM algorithm is an iterative algorithm to find MAP estimates or the maximum likelihood estimates. It is commonly used when some necessary data to establish the likelihood function are missing. Instead, the algorithm maximizes the expectation, with respect to the "missing" data, of the likelihood function.

@@

* E-step: to calculate the expectation of the likelihood function conditioning on some "missing" data
* M-step: to maximize the expected likelihood function to calculate the parameter of interest

>>>>

Following the GLM notation introduced in Section \ref{sec:intro}, we have a generalized additive model and its matrix form,
\begin{equation}\label{eq:gam}
E(Y|\bs x) = g^{-1}(a + \sum\limits^p_{j=1}B_j(x_j)) = g^{-1}(a + \sum\limits^p_{j=1} \bs \beta_j^T \bs X_j),
\end{equation}
and the data distribution is
\begin{equation}
f(\bs y| \bs \beta, \phi) = \prod\limits^n_{i=1}f(y_i|\bs \beta, \phi),\nonumber
\end{equation}
with smoothing functions $B_j(x_j)$ of the variable $x_j, j = 1, \dots, p.$ The basis function matrix, i.e. the design matrix derived from the smoothing function $B_j(x_j)$, is denoted $\bs X_j$ for the variable $x_j$. The dimension of the design matrix depends on the choice of the smoothing function. For example, a cubic spline function with $k$ knots produces a $k+3$ elements column vector. Its corresponding smoothing penalty matrix is denoted as $\bs S_j$. $\bs \beta_j$ denotes the basis function coefficients for the $j$th variable such that $B_j(x_j) = \bs \beta_j^T \bs X_j$. With slight abuse of notation, we denote vectors and matrices in bold fonts $\bs \beta, \bs X$ with conformable dimensions, where scalar and random variables are denoted in with unbold fonts $\beta, X$. The matrix transposing operation is denoted with a superscript $^T$. To note, the proposed model can include parametric forms of variables in the model, treating the parametric function as a special case of the smoothing function, e.g. $B_j(x_j) = x_j$ with the smoothing penalty matrix defined as $\bs S_j = \begin{bmatrix}0\end{bmatrix}$.

To encourage proper smoothing of the functions, we adopt the idea of smoothing penalties from smoothing spline models. The basic idea is to set up a smoothing penalty, described in Equation (\ref{eq:smoothpen}), in the prior density function. However, the direct integration of smoothing penalty with sparsity penalty is not obvious. Marra and  Wood [@Marra2011] proposed a re-parameterization procedure to accommodate the smoothing penalty implicitly. Given the smoothing penalty matrix $\bs S_j$ is symmetric and positive semi-definite for univariate smoothing functions, we apply eigendecomposition on the penalty matrix $\bs S = \bs U \bs D \bs U^T$ , where the matrix $\bs D$ is diagonal with the eigenvalues arranged in the ascending order. To note, $\bs D$ can contain elements of zeros on the diagonal, where the zeros are associated with the linear space of the smoothing function. For the most popular smoothing function, cubic splines, the dimension of the linear space is one. Hereafter, we focus on discussing a uni-dimensional linear space for simplicity; however, it generalizes easily to the cases where the linear space is multidimensional. We further write the orthonormal matrix $\bs U \equiv \begin{bmatrix} \bs U^0 : \bs U^{\tp}\end{bmatrix}$ containing the eigenvectors as columns in the corresponding order to $\bs D$. That is, $\bs U$ contains the eigenvectors $U^0$ with zero eigenvalues for the linear space and $\bs U^{\tp}$ contains the eigenvectors (as columns) for the non-zero eigenvalues, i.e. the non-linear space. We multiply the basis function matrix $\bs X$ with the orthonormal matrix $\bs U$ for the new design matrix ${\bs X}^\repa = \bs X \bs U \equiv \begin{bmatrix} X^0 : \bs X^{\tp} \end{bmatrix}$. An additional scaling step is imposed on $\bs X^{\tp}$ by the non-zero eigenvalues of $\bs D$ such that the new basis function matrix $\bs X^\ast$ can receive uniform penalty on each of its dimensions. With slight abuse of the notation, we drop the superscript $^\repa$ and denote $\bs X_j \equiv \begin{bmatrix} X_j^0 : \bs X_j^{\tp} \end{bmatrix}$ as the basis function matrix for the $j$th variable after the re-parameterization. A spline function can be expressed in the matrix form
$$
B_j(x_j) = B_j^0(x_j) + B_j^\tp (x_j) = \beta_j X^0_j + \bs {\beta_j^\tp}^T \bs X_j^\tp ,
$$
and the generalized additive model in Equation (\ref{eq:gam}) be
\begin{equation}\label{eq:gam-repa}
E(Y|\bs x) = g^{-1}(a + \sum\limits^p_{j=1} B_j(x_j)) = g^{-1}(a + \sum\limits^p_{j=1} \bs \beta_j^T \bs X_j) = g^{-1}\left[a + \sum\limits^p_{j=1} (\beta_j X^0_j + {\bs \beta_j^\tp}^T \bs X_j^\tp)\right],
\end{equation}
where the coefficients $\bs \beta_j \equiv \begin{bmatrix} \beta_j : \bs \beta^\tp_j \end{bmatrix}$ is an augmentation of the coefficient scalar $\beta_j$ of linear space and the coefficient vector $\bs \beta^\tp_j$ of non-linear space.

@@

The major criticism of early spike-and-slab models is being computationally prohibitive. [@Bai2021Review] Since then, many studies focus on alleviating the computational burden that sampling algorithms bear, which include EMVS [@Rockova2014a] and spike-and-slab Lasso [@Rockova2018b; @Rockova2018]. Particularly, the development of spike-and-slab Lasso  substantially improves the scalability of SS models, setting up the theoretical foundation for generalized models in -omics data analysis [@Tang2017a; @Tang2017; @Tang2018; @Tang2019]. The spike-and-slab Lasso (SSL) prior is composed of two double exponential distributions with mean 0 and different dispersion parameters, $0 < s_0 < s_1$, mathematically,
\begin{equation} 
\beta | \gamma \sim (1-\gamma)DE(0, s_0) + \gamma DE(0, s_1), 0 < s_0 < s_1.\nonumber
\end{equation}
Given that both double exponential distributions have a mean of 0 and the latent indicator $\gamma$ can only take the value of 0 or 1, the mixture double exponential distribution can be formulated as one single double exponential density,
\begin{equation} \label{eq:ssl}
\beta | \gamma \sim DE(0, S), 0 < s_0 < s_1,
\end{equation}
with the scale parameter $S = (1-\gamma)s_0 + \gamma s_1$. The SSL also mitigates the problem of EMVS where the weak signals are not shrink to zero, and hence is preferred in high-dimensional data analysis. We notice that Bai [@Bai2021] is the first to apply spike-and-slab lasso prior in the GAM framework, where the densities of the spike and slab components take the group lasso density [@Xu2015]. Bai [@Bai2021] takes an "all-in-all-out" strategy for function selection.

### Spike-and-Slab Lasso Prior

We introduce a novel prior for GAMs, particularly for high-dimensional nonlinear modeling with bi-level selection. The proposed prior extends from the spike-and-slab lasso prior described in Equation (\ref{eq:ssl}). Given the re-parameterized deign matrix $\bs X_j = \begin{bmatrix} X^0_j : \bs X_j^\tp \end{bmatrix}$ for the $j$th variable, we impose a two-part SSL prior to the coefficients $\bs \beta_j = \begin{bmatrix} \beta_j : \bs \beta_j^\tp \end{bmatrix}$. Specifically, we impose independent group priors on the linear space coefficients and on the nonlinear space coefficients respectively, 
\begin{align}\label{eq:bham_ssl}
  \beta_{j} |\gamma_{j},s_0,s_1 &\sim DE(0,(1-\gamma_{j}) s_0 + \gamma_{j} s_1) \nonumber \\
  \beta^\tp_{jk} | \gamma^\tp_{j},s_0,s_1 &\simiid DE(0,(1-\gamma^\tp_{j}) s_0 + \gamma^\tp_{j} s_1), k=1,\dots, K_j
\end{align}
where $\gamma_{j}\in\{0,1\}$ and $\gamma^\tp_{j}\in \{0,1\}$ are two latent indicator variables, indicating if the model includes the linear effect and the nonlinear effect of the $j$th variable respectively. $s_0$ and $s_1$ are scale parameters, assuming $0 < s_0 < s_1$ and given. These scale parameters $s_0$ and $s_1$ can be treated as tuning parameters and optimized via cross-validation. A discussion of how to choose the scale parameters comes in Section \ref{sec:tune}. To note, this prior differs from previous group spike-and-slab lasso priors [@Tang2018; @Tang2019], as the $\beta_j$ and $\bs \beta^\tp_j$ have different indicator variables $\gamma_j$, $\gamma_j^\tp$ respectively. It is possible to add a more restrictive assumption on the priors, assuming that one indicator variable decides the inclusion of both the linear effect and nonlinear effect, i.e. $\gamma_j = \gamma^\tp_j$.  This converges to the SB-GAM [@Bai2021]. Conversely, it is also possible to relax the assumption such that each coefficient $\beta_{jk} \in \bs \beta^\tp_j$ has its own latent indicator $\gamma_{jk}$, but at the cost of complicating the bi-level function selection. This reduces the proposed prior to the classic spike-and-slab lasso prior.

@@

The rest of the hierarchical prior follows the traditional spike-and-slab lasso prior: we set up hyper-priors on $\gamma_j, \gamma^\tp_j$ to allow local adaption of the shrinkage using a Bernoulli distribution, written as binomial distribution of one trial. The two indicators of the $j$th predictor, $\gamma_j$ and $\gamma^\tp_j$, shares the same probability parameter $\theta_j$,
$$
\begin{aligned}
&\gamma_{j} | \theta_j \sim Bin(1, \theta_j) & & 
&\gamma_{j}^\tp | \theta_j \sim Bin(1, \theta_j).
\end{aligned}
$$
This is to leverage the fact that the probability of selecting the bases of a smoothing function should be similar, while allowing different penalty on the linear space and non-linear space of the smoothing function. The hyper prior of $\gamma_{j}$ decides the sparsity of the model at the function selection level, while that of $\gamma_{j}^\tp$ decides the smoothness of the spline function at basis function level. Meanwhile, we specify that $\gamma_{j}$ and $\gamma_{j}^\tp$ are independently distributed for analytic simplicity. We further specify the parameter $\theta_j$ follows a beta distribution with given shape parameters $a$ and $b$,
$$
\theta_j \sim Beta(a, b).
$$
The beta distribution is a conjugate prior for the binomial distribution and hence provides some computation convenience. For simplicity, we focus on a special case of beta distribution, uniform (0,1), i.e. $a = 1, b = 1$. When the variable have large effects in any of the bases, the parameter $\theta_j$ will be estimated large, which in turn encourages the model to include the rest of bases, achieving the local adaption among spline bases. Hereafter, we refer Bayesian hierarchical generalized additive models with the spike-and-slab lasso prior as the BHAM-SSL, and visually presented in Figure \ref{fig:SSprior}.

@@

\matrix (linearPrior) [matrix of nodes, column sep = 0mm, row sep = 0.7cm] {
  \node (linearGamma) [outCome] { $\gamma_j \sim Bin(1, \theta_j) $ };\\
  \node (linearBeta) [outCome] { $\beta_j \sim DE(0,(1-\gamma_{j}) s_0 + \gamma_{j} s_1)$};\\
};
\matrix (penPrior) [right = 2cm of linearPrior, matrix of nodes, column sep = 0mm, row sep = 0.7cm] {
  \node (penGamma) [outCome] { $\gamma_{j}^\tp \sim Bin(1, \theta_j)$ };\\
  \node (penBeta) [outCome] { $\beta_{jk}^\tp \sim  DE(0,(1-\gamma^\tp_{j}) s_0 + \gamma^\tp_{j} s_1)$};\\
};

\node (s) [staticCompo]  at ($(linearBeta)!0.5!(penBeta)$)  {($s_0, s_1$)};
\node (Beta) [staticCompo, below = 1cm of s] {$\bs \beta = (\beta_1, \bs \beta^\tp_1, \dots,\beta_j, \bs \beta^\tp_j , \dots,\beta_p, \bs \beta^\tp_p) $};
\node (Theta)[outCome, above = 2cm of s] {$\theta_{j} \sim Beta(a, b)$};
\node (ab)[staticCompo, above = 0.5cm of Theta] {$(a, b)$};
\node (Y) [outCome, below = 1cm of Beta] {$y_i \sim Expo. Fam. (g^{-1}(\bs \beta^T \bs X_i), \phi)$};

@@

With the re-parameterization step of the basis function matrix $\bs X$, it is possible to generalized the SSL prior to other priors, for example normal priors for ridge-type regularization and mixture normal prior for spike-and-slab regularization. These priors would work better in low and medium dimensional settings where the sparse assumption is not necessary. Here we elaborate the mixture normal prior as a demonstration of applying continuous spike-and-slab prior in BHAM.

A spike-and-slab mixture normal spline prior can be expressed as 
\begin{align*}
  \beta_{j} |\gamma_{j},s_0,s_1 &\sim N(0,(1-\gamma_{j}) s_0 + \gamma_{j} s_1)\\
  \beta^\tp_{jk} | \gamma^\tp_{j},s_0,s_1 &\simiid N(0,(1-\gamma^\tp_{j}) s_0 + \gamma^\tp_{j} s_1), k=1,\dots, K_j.
\end{align*}
Similar to the BHAM-SSL prior in Equation (\ref{eq:bham_ssl}), $0 < s_0 < s_1$ are tuning parameters and can be optimized via cross-validation. One of the critics received by the spike-and-slab mixture normal prior is that the tails of a normal distribution diminishes to zero too fast, which causes problems when estimating the large effects. Distributions with heavier tails can be used as an alternative, for example mixture Student's $t$ distribution with small degree of freedom.

@@

The proposed models can be fitted with MCMC algorithms. Nevertheless, the computational burden of MCMC algorithms creates scalability issues. George and McCulloch[@George1997] examined the computation speed for various MCMC algorithms with spike-and-slab mixture normal priors, and suggested MCMC algorithms works well for medium size ($p$=25) of predictors with only linear effects. However, it is not feasible for high-dimensional data analysis where the number of predictors easily exceeds 100. Specific to additive models, each predictor would expand to multiple new "predictors" via basis functions, creating greater computational demands. Scheipl et al. [@Scheipl2013] demonstrated the computational demands of a MCMC algorithm for fitting spike-and-slab GAM grow exponentially as $p$ increases modestly via simulation studies. Hence, we feel compelled to develop scalable algorithms for fitting Bayesian hierarchical additive models in high-dimensional settings.

As alternatives to the sampling algorithms for fitting Bayesian models, optimization algorithms focus on the maximum a posteriori (MAP) estimates and speed up the model fitting process at the cost of uncertainty measures. The earlier work for fitting spike-and-slab models using optimization algorithms includes EMVS [@Rockova2014a]. Rockova and George[@Rockova2014a] proposed an expectation-maximization (EM) based algorithm to fit models that use continuous mixture normal priors. 
<!-- The coefficient $\beta_j$ independently follows a mixture normal distribution $N(0, \sigma^2((1-\gamma_i)v_0 + \gamma_i v_1)$ for $0 \leq v_0 < v_1$. The parameter $\sigma^2$ follows a inverse gamma prior, $IG(v/2, v\lambda /2)$. Like in the spike-and slab family, the latent binary variables $\bs \gamma$ follows a Bernoulli prior with a hyper prior $\theta$ follows a beta distribution, or simpler the uniform (0,1).  -->
In the E step, the latent binary indicators $\gamma$s are treated as the missing data, and the posterior means are calculated conditioning on the current value of other parameters; in the M step, a ridge estimator was used to update the coefficients, followed by updating $\phi, \theta$. 
<!-- As the ridge regression creates small but nonzero coefficients, a further step is needed to select the variable: EMVS further suggests to threshold the inclusion probability for variable selection or generate regularization plot with different values of the spike scale parameter. EMVS is also compatible with structured prior when dealing with grouped variables. The group structure can be imposed on the prior of the latent binary variable $\gamma$ via either logistic regression or markov random field.  -->
The same authors [@Rockova2018b; @Rockova2018] further combined the EM algorithm with coordinate descent algorithm to fit SSL models.
<!-- the continuous spike-and-slab Gaussian prior to the spike-and-slab double exponential distribution, as known as spike-and-slab lasso (SSL). The authors also lay the theoretical discussion of comparing spike-and-slab from the penalty perspective and the Bayesian perspective. The SSL also mitigated the problem of EMVS that the weak signals are not shrink to zero. -->
Yi and his group independently developed the EM-Iterative Weighted Least Square and EM-cyclic coordinate descent algorithms to fit models with broader class of priors, SSL included.[@Yi2019] These algorithms were implemented for generalized linear models[@Tang2017a], Cox proportional hazards models [@Tang2017] and their grouped counterparts[@Tang2018; @Tang2019]. Both EM based algorithms provide deterministic solutions, which becomes a popular property for reproducible research

@@

In this section, we extend the two EM-based algorithms, EM-CD and EM-IWLS algorithms, to fit BHAMs. To note, the two proposed algorithms provides different utilities. The EM-CD algorithm is specifically for fitting BHAM-SSL with an expedited performance, recommending to use in high and ultra-high dimensional setting. A specific concern of EM-CD algorithm is that it provides no information for inference. In contrast, the EM-IWLS can extract the variance-covariance matrix and be used to construct Wald type inference. Moreover, the EM-IWLS is a more general model fitting algorithm that can be used for fitting not only SSL and continuous SS priors but also Student's t-priors and double exponential priors. To note, SB-GAM[@Bai2020; @Bai2021] also used an EM-CD algorithm. The main difference between proposed EM-CD algorithm and that in SB-GAM is that SB-GAM uses a blocked CD algorithm for their group prior, while the proposed prior is pairwise independent requiring no special treatment in the CD algorithm.

@@

EM algorithm is an iterative algorithm to find MAP estimates or the maximum likelihood estimates. It is commonly used when some necessary data to establish the likelihood function are missing. Instead of maximizing the the likelihood function, the algorithm maximizes the expectation of the likelihood function with respect to the "missing" data.

@@

* E-step: to calculate the expectation of the posterior density function with respect to some "missing" data
* M-step: to maximize the expectation derived in the E-step to calculate the parameter of interest

For BHAMs, we define the parameters of interest as $\Theta = \{\bs \beta, \bs \theta, \phi\}$ and consider the latent binary indicators $\bs \gamma$ as nuisance parameters of the model, in other words the "missing" data. Our objective is to find the parameters $\Theta$ that maximize the posterior density function, or equivalently, the logarithm of the density function, 
$$
\begin{aligned}
& \text{argmax}_{\Theta}
\log f(\Theta, \bs \gamma| \textbf{y}, \textbf{X}) \\
&= \log f(\textbf{y}|\bs \beta, \phi) + \sum\limits_{j=1}^p\left[\log f(\beta_j|\gamma_j)+\sum\limits_{k=1}^{K_j} \log f(\beta^{\tp}_{jk}|\gamma^{\tp}_{j})\right]\\
& +\sum\limits_{j=1}^{p} \left[ (\gamma_j+\gamma_{j}^{\tp})\log \theta_j + (2-\gamma_j-\gamma_{j}^{\tp}) \log (1-\theta_j)\right] +  \sum\limits_{j=1}^{p}\log f(\theta_j),
\end{aligned}
$$  
where $f(\textbf{y}|\bs \beta, \phi)$ is the data distribution and $f(\theta)$ is the Beta(1,1) density.
